Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-05 manuscript_v2.tex
updated files

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information (Chafe,
1976). The Givenness Hierarchy is akin to other previously proposed
models of reference control within the topicality and saliency paradigm
such as the Accessibility Hierarchy (Ariel, 1990), Topic Continuity
(Givon, 1983), and Familiarity Scale (Prince, 1981). Gundel et al.
(1993) argued that the distribution of forms is not random but abides by
the Gricean Maxim of Quantity (Grice, 1975). That is to say, individuals
must make the most informative and succinct choice possible to retain
economy of form. When multiple expressions are available for the same
referent as in an utterance like (1), phonetically simplest form is
selected with the least redundant information are selected based on an
economy principle. In (1), REs for ``Eddie'' are given in bold. All
three REs (nominal, pronominal, and null marking) refer to the same
entity; however, the quantity of marking decreases as the referent's
accessibility increases. On this ground, for highly accessible referents
(i.e., those that are highly active in the mind), linguistic forms with
the least phonetic content are selected for the least amount of
processing effort.

\begin{exe}
    \ex
    So \textbf{Eddie} turned around. \hfill Williams (1988; p. 343, ex. 7) \\
\textbf{He} said, "youse got a problem?" \\
"Yeah, we want you," they say.
So-\textbf{$\emptyset$} walked right up to them \\
\end{exe}

@@

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives. Building on Gundel et al.
(1993)'s work, Swabey (2002) demonstrated that, like other spoken
languages such as Russian, Chinese, and Japanese, American Sign Language
(ASL) signers use bare nouns for introduced and reintroduced discourse
contexts whereas null marking is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and were thus economic in their reference tracking.
Similar findings have been reported for German Sign Language (DGS)
(Perniss and Özyürek, 2015) and British Sign Language (BSL) (Morgan,
2005). It is important to note, however, that these research either
collapsed native and late signers into a single group and labeled them
as ``native-like'' or explicitly compared native deaf signers with late
deaf signers (with language deprivation) in terms of narrative
competence.

@@

Commonly in both spoken and sign languages nouns and noun phrases are
used as referring expressions. Spoken languages use pronouns, determines
and demonstratives where sign languages use a pointing sign, usually
glossed as IX, which assumes similar functions (Lillo-Martin and Klima,
1990). In addition, the pointing sign is used to referentially anchor a
referent to a location in signing space. For instance, when the signer
introduces \textsc{elvan} for the first time, s/he can assign her to the
right. This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (2a) there are two
referents, \textsc{elvan}, who is referentially-assigned to the right
side of the signer, represented with the ``a'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``b'' subscript. When the signer makes a
contrast in (2b), as a pronominal use, she points to the right side to
refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represents the referential locus assigned to the referents
in (3).

\begin{exe}
    \ex 
\begin{xlist}
    \ex
    \textsc{lab place elvan ix\textsubscript{a} ten years work. demet ix\textsubscript{b} last year start.} \\
    ‘In the lab, Elvan has been working for 10 years. Demet started last year.’
    \\
    \ex
    \textsc{ix\textsubscript{a} e-l-a-n know. ix\textsubscript{b} learn.}\\
    ‘She (Elvan) knows ELAN (the annotation program). She (Demet) is learning.’
    \end{xlist}
    \end{exe}

    \begin{exe}
    \ex Signing space, signer, example referential loci for the referents \textsc{elvan} (a) and \textsc{demet} (b) \\

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). Classifiers denote sets of objects based on semantic
criteria such as size, shape, body-part, texture and sometimes function.
Usually, the verbal classifier and the cross-referenced noun co-occur in
the same clause when the argument is introduced into discourse. The
classifier can track the argument on its own on subsequent occurrences.
In sign languages, a verbal classifier morpheme is encoded by a
\emph{handshape} that is articulated phonologically co-temporally with a
movement, which encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). These two infused morphemes form a complex
predicate. Some classifier handshapes and the set of objects that they
denote are below. Some classifier handshapes and the set of objects that
they denote are presented below in (4).

\begin{exe}
\ex 
\includegraphics[width=25pt]{pictures/flathand.png}: flat entities, surfaces, vehicles.\\
\includegraphics[width=25pt]{pictures/flat.png}: human agent holding a flat entity.\\
\includegraphics[width=25pt]{pictures/small_legged.png}: small animal legs.\\
\includegraphics[width=25pt]{pictures/index.png}: thin entities (like a pencil); upright beings.\\
\includegraphics[width=25pt]{pictures/hold.png}: human agent holding a thin entity (like a pencil or a branch of flower). 
\end{exe}

In (5), in the first clause, a referential locus, i.e.~locus-a, is
assigned to the locative source argument \textsc{cage}. In the second
clause, the core argument \textsc{bird} is cross-referenced with the
classifier handshape, i.e.~small-legged animal. It flies from the
referential locus of the \textsc{cage} (locus-a) to the referential
locus of the goal \textsc{tree} (locus-b).

\begin{exe}
    \ex 
\begin{xlist}
    \ex \textsc{cage} assigned to the referential locus-a. \\ \\
    \textsc{garden bird cage-a open} \\
    ‘In the garden, there is a cage and it is open.’
    \\

@@

\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex Noun modified by index (IX)\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ IX \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ IX \textsc{sick} \\ \\
\z \z
\begin{multicols}{4}
\begin{exe}
\ex
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

\ex Plain verb  \\\glll
\includegraphics[width=50pt]{pictures/4a_2.png}  \\ \textsc{enter}\\ \\

\ex Agreement verb  \\\glll 
\includegraphics[width=50pt]{pictures/4b_1.png} 
\includegraphics[width=50pt]{pictures/4b_2.png} \\ {\textsc{$_{1}$hit$_{3}$}} {} \\ \\ 
\end{xlist}
\end{exe}
\end{multicols}
\footnotetext{We use small capitals to indicate sign glosses.}

@@

In TİD and many other sign languages, it is possible to use bare noun
phrases (\textsc{bare nom}) as in (1a) as well as fingerspelled nouns
(\textsc{fs nom}) as in (1c). Fingerspelling refers to the use of the
manual alphabet which is derived or borrowed from a surrounding written
language (Kubuş, 2008). Nouns can also be modified (\textsc{mod nom}) by
classifiers as in (1b) (Frederiksen and Mayberry, 2016) or by a
prenominal or postnominal IX functioning as a determiner or a
demonstrative as in (1d) (Maclaughlin, 1997).

@@

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse (Emmorey, 1996). The same IX function has also been identified
for TİD (Sevinç, 2006). The example in (2) illustrates a
spatially-anchored pronominal reference for topic re-introduction. These
signs are thus rarely used for the first mention of an entity (Perniss
and Özyürek, 2015). Other interpretations consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns (Koulidobrova and Lillo-Martin, 2016). It is important to
mention here that in some spoken languages (e.g., German and Dutch),
demonstratives can also refer to persons (Bosch and Umbach, 2007) and
behave similarly to personal pronouns. However, for the purposes of the
present study, we will refer to IX constructions as pronominal IX and
particularly examine their non-deictic or personal pronoun realizations
as a possible discourse strategy to mainly mark topic shift among native
and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also common and found in TİD (see Dikyuva
et al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL) occur with transitive verbs
and depict how an object is handled or manipulated by agent. On this
ground, an HCL\footnote{As discussed before, an HCL is transitive in
  nature and depicts the handling action performed by ananimate agent.
  This is why most research tend to consider HCLs as an extension to
  constructed action (Cormier et al., 2013; Dudis, 2004).} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (3).

@@

Parallel to spoken languages, many sign languages have been reported to
allow null subjects and/or objects and make use of null arguments when
referring to highly accessible referents which are still active in the
mental representation of the addressee. In signed discourse, this can be
achieved by employing certain verbal constructions. To identify the
predicate types in ASL, Padden (1986) suggested a triadic categorization
of verbs in sign languages: plain verbs, agreement verbs, and spatial
verbs. According to this classification, plain verbs (plain) (4b) do not
agree with the subject or the object. This means that they are not
inflected for number or person. This contrasts with agreeing verbs
(agreement) (4c) the starting or the ending point of which can be
modified to specific loci for person and number agreement while spatial
verbs (as in \textsc{ix book put} `I placed the book there') agree with
locative arguments but like plain verbs, they are not inflected for
number or person. TİD too licenses null arguments. Following this
classification, Sevinç (2006) argued that although it is common in TİD
to drop arguments, there were certain limitations on the null marking of
the arguments for agreement and plain verbs.

@@

In both communication modalities (visual and auditory), children start
to produce more complex and coherent stories with appropriate and
informative linguistic forms to indicate anaphoric or cataphoric
references after age 10 (Hickmann et al., 1996; Rathmann et al., 2007).
Prior to this age, both signing and speaking children fail to refer to
entities consistently since their pragmatic abilities are yet to
develop. The manipulation of linguistic forms for discourse functions to
signal topic continuance or shift is difficult to master until the
beginning of formal education. These findings are also comparable across
different sign languages including British Sign Language (BSL) (Morgan,
2005), ASL (Loew, 1984), and German Sign Language (Deutsche
Gebärdensprache - DGS) (Becker, 2009), pointing out to the fact that the
mastery of reference tracking in narratives is one of the most
protracted achievements in the acquisition of signed discourse.

@@

Still, only very few research examined whether linguistically deprived
signers (i.e., deaf signers with delayed exposure to a sign language)
also display redundancy in their reference tracking. Becker (2009)
examined 6 DGS signing deaf children (3 native and 3 late)`s narrative
abilities. She reported that late signing children generated more
explicit forms such as nominals when referring to the characters in the
story compared to native learners, concluding that first language
deprivation results in an inadequate representation of the addressee and
less efficient use of morphosyntax. Comparably, Cormier et al. (2013)
investigated the use of constructed action in 15 BSL among deaf native
and late adult signers and found that late signers were redundant in
topic continuity. In other words, the late signers produced more
explicit forms like nominals, and fewer instances of null marking when
maintaining previous referents. Most relevantly to the present study,
Gür (2018) studied the development of narrative skills among 10 native
and 10 late deaf children and deaf adult signers of TİD. She found that
native-signing children and adults had more instances of explicit
referent introduction, set the scene more frequently, and employed
constructed action in an appropriate manner more than corresponding
late-signing children and adults. Although this study presented
preliminary findings on age of acquisition effects on narrative
competence, it does not investigate how native and late signers employ
different linguistic tools based on the accessibility of the referent.
Therefore, to our knowledge, there is a lack of research directly
examining the effects of first language deprivation on deaf adult
signers' selection of REs in signed narratives.

@@

Following Berman and Slobin (2013), we first identified the clause
boundaries in each sentence indicated by the presence of predicates and
informative non-manual markers (e.g., head nod). On the second tier, we
coded the discourse status for each referent. Having adapted the
classification of Gullberg (2006), we considered all referents mentioned
for the first time as introduced regardless of their syntactic function
in the sentence. Only subjects were coded for maintenance and
re-introduction. We coded subject referents as maintained only if they
had been referred to in the previous clause. To this end, the referents
which had been discontinued across one or more sentences were all
regarded re-introduced into the discourse. On a third tier, we had the
RE type for a single reference with the main categories nominal,
pronominal, null marking, and classifier. Under nominal, we identified
bare nouns, modified nouns, and fingerspelled nouns. For classifiers, we
distinguished among three types of constructions: WECL, BPCL, and ExtCL.
Finally, we had CA, agreement, plain tags for null constructions. We
were unable to observe any instance of HCLs occurring in isolation
(i.e., not part of constructed action). The annotation tiers are
presented in Table \ref{tab:tab-2}.

@@

\begin{table}
\small
\caption{Distribution of all discourse status by referring expression types}
\label{tab:my-table}
\begin{tabular}{llll}
\hline
                     & Introduced    & Maintained     & Re-introduced \\ \hline
\textit{Nominal:}    & 92.78\% (989) & 5.63\% (93)    & 63.65\% (907) \\
\textit{Pronominal:} & 0.09\% (1)    & 0.24\% (4)     & 2.46\% (35)   \\
\textit{Classifier:} & 5.35\% (57)   & 18.29\% (302)  & 8\% (114)     \\
\textit{Null:}       & 1.78\% (19)   & 75.83\% (1252) & 25.89\% (369) \\ \hline
\end{tabular}%
\end{table}

@@

\begin{table}
\small
\caption{Distribution of nominal referring expressions by discourse status}
\label{tab:my-table2}
\begin{tabular}{lllll}
\hline
                        & Bare Noun  & Modified Noun & FS Noun \\ \hline
\textit{Introduced:}    & 48.93\% (842) & 53.56\% (139)     & 100\% (8) \\
\textit{Maintained:}    & 4.82\% (83)  & 3.85\% (10)      & 0\% (0) \\
\textit{Re-introduced:} & 46.25\% (796) & 42.69\% (111)     & 0\% (0) \\ \hline
\end{tabular}
\end{table}

>>>>

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information (Chafe,
1976). The Givenness Hierarchy is akin to other previously proposed
models of reference control within the topicality and saliency paradigm
such as the Accessibility Hierarchy (Ariel, 1990), Topic Continuity
(Givon, 1983), and Familiarity Scale (Prince, 1981). Gundel et al.
(1993) argued that the distribution of forms is not random but abides by
the Gricean Maxim of Quantity (Grice, 1975). That is to say, individuals
must make the most informative and succinct choice possible to retain
economy of form. When multiple expressions are available for the same
referent as in an utterance like (\ref{ex:1}), phonetically simplest
form is selected with the least redundant information are selected based
on an economy principle. In the utterance, REs for ``Eddie'' are given
in bold. All three REs (nominal, pronominal, and null marking) refer to
the same entity; however, the quantity of marking decreases as the
referent's accessibility increases. On this ground, for highly
accessible referents (i.e., those that are highly active in the mind),
linguistic forms with the least phonetic content are selected for the
least amount of processing effort.

\begin{exe}
    \ex \label{ex:1}
    So \textbf{Eddie} turned around. \hfill Williams (1988; p. 343, ex. 7) \\
\textbf{He} said, "youse got a problem?" \\
"Yeah, we want you," they say.
So-$\pmb{\emptyset}$ walked right up to them \\
\end{exe}

@@

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives. Building on Gundel et al.
(1993)'s work, Swabey (2002) demonstrated that, like other spoken
languages such as Russian, Chinese, and Japanese, American Sign Language
(ASL) signers use bare nouns for introduced and re-introduced discourse
contexts whereas null marking is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and were thus economic in their reference tracking.
Similar findings have been reported for German Sign Language (Deutsche
Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British Sign
Language (BSL) (Morgan, 2005). It is important to note, however, that
these research either collapsed native and late signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with language deprivation) in terms
of narrative competence.

@@

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions (Lillo-Martin and Klima, 1990).
In addition, the pointing sign is used to referentially anchor a
referent to a location in signing space. For instance, when the signer
introduces \textsc{elvan} for the first time, s/he can assign her to the
right. This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``a'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``b'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represents the referential locus assigned to the referents
in (\ref{ex:3}).

\begin{exe}
    \ex 
\begin{xlist}
    \ex \label{ex:2a}
    \textsc{lab place elvan ix\textsubscript{a} ten years work. demet ix\textsubscript{b} last year start.} \\
    ‘In the lab, Elvan has been working for 10 years. Demet started last year.’
    \\
    \ex \label{ex:2b}
    \textsc{ix\textsubscript{a} e-l-a-n know. ix\textsubscript{b} learn.}\\
    ‘She (Elvan) knows ELAN (the annotation program). She (Demet) is learning.’
    \end{xlist}
    \end{exe}

    \begin{exe}
    \ex \label{ex:3} Signing space, signer, example referential loci for the referents \textsc{elvan} (a) and \textsc{demet} (b) \\
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). Classifiers denote sets of objects based on semantic
criteria such as size, shape, body-part, texture and sometimes function.
Usually, the verbal classifier and the cross-referenced noun co-occur in
the same clause when the argument is introduced into discourse. The
classifier can track the argument on its own on subsequent occurrences.
In sign languages, a verbal classifier morpheme is encoded by a
\emph{handshape} that is articulated phonologically co-temporally with a
movement, which encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). These two infused morphemes form a complex
predicate. Some classifier handshapes and the set of objects that they
denote are below. Some classifier handshapes and the set of objects that
they denote are presented below in (\ref{ex:4}).

\begin{exe}
\ex \label{ex:4}
\includegraphics[width=25pt]{pictures/flathand.png}: flat entities, surfaces, vehicles.\\
\includegraphics[width=25pt]{pictures/flat.png}: human agent holding a flat entity.\\
\includegraphics[width=25pt]{pictures/small_legged.png}: small animal legs.\\
\includegraphics[width=25pt]{pictures/index.png}: thin entities (like a pencil); upright beings.\\
\includegraphics[width=25pt]{pictures/hold.png}: human agent holding a thin entity (like a pencil or a branch of flower). 
\end{exe}

In (\ref{ex:5}), in the first clause, a referential locus, i.e.~locus-a,
is assigned to the locative source argument \textsc{cage}. In the second
clause, the core argument \textsc{bird} is cross-referenced with the
classifier handshape, i.e.~small-legged animal. It flies from the
referential locus of the \textsc{cage} (locus-a) to the referential
locus of the goal \textsc{tree} (locus-b).

\begin{exe}
    \ex \label{ex:5}
\begin{xlist}
    \ex \textsc{cage} assigned to the referential locus-a. \\ \\
    \textsc{garden bird cage-a open} \\
    ‘In the garden, there is a cage and it is open.’
    \\

@@

\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:6a} Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex \label{ex:6b} Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex \label{ex:6c} Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex \label{ex:6d} Noun modified by index (\textsc{IX})\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ \textsc{ix} \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea \label{ex:7} Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ \textsc{ix} \textsc{sick} \\ \\
\z \z
\begin{multicols}{4}
\begin{exe}
\ex \label{ex:8}
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:9a} Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

\ex \label{ex:9b} Plain verb  \\\glll
\includegraphics[width=50pt]{pictures/4a_2.png}  \\ \textsc{enter}\\ \\

\ex \label{ex:9c} Agreement verb  \\\glll 
\includegraphics[width=50pt]{pictures/4b_1.png} 
\includegraphics[width=50pt]{pictures/4b_2.png} \\ {\textsc{$_{1}$hit$_{3}$}} {} \\ \\ 
\end{xlist}
\end{exe}
\end{multicols}
\footnotetext{We use small capitals to indicate sign glosses.}

@@

In TİD and many other sign languages, it is possible to use bare noun
phrases as in (\ref{ex:6a}) as well as fingerspelled nouns as in
(\ref{ex:6c}). Fingerspelling refers to the use of the manual alphabet
which is derived or borrowed from a surrounding written language (Kubuş,
2008). Nouns can also be modified by classifiers as in (\ref{ex:6b})
(Frederiksen and Mayberry, 2016) or by a prenominal or postnominal IX
functioning as a determiner or a demonstrative as in (\ref{ex:6d})
(Maclaughlin, 1997).

@@

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse (Emmorey, 1996). The same IX function has also been identified
for TİD (Sevinç, 2006). The example in (\ref{ex:7}) illustrates a
spatially-anchored pronominal reference for topic re-introduction. These
signs are thus rarely used for the first mention of an entity (Perniss
and Özyürek, 2015). Other interpretations consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns (Koulidobrova and Lillo-Martin, 2016). It is important to
mention here that in some spoken languages (e.g., German and Dutch),
demonstratives can also refer to persons (Bosch and Umbach, 2007) and
behave similarly to personal pronouns. However, for the purposes of the
present study, we will refer to IX constructions as pronominal IX and
particularly examine their non-deictic or personal pronoun realizations
as a possible discourse strategy to mainly mark topic shift among native
and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also common and found in TİD (see Dikyuva
et al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL) occur with transitive verbs
and depict how an object is handled or manipulated by agent. On this
ground, an HCL\footnote{As discussed before, an HCL is transitive in
  nature and depicts the handling action performed by ananimate agent.
  This is why most research tend to consider HCLs as an extension to
  constructed action (Cormier et al., 2013; Dudis, 2004).} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

@@

Parallel to spoken languages, many sign languages have been reported to
allow null subjects and/or objects and make use of null arguments when
referring to highly accessible referents which are still active in the
mental representation of the addressee. In signed discourse, this can be
achieved by employing certain verbal constructions or constructed action
(CA). The latter construction exemplified in (\ref{ex:9a}) refers to
cases in which the signers engage with the space in front of themselves
as if they are interacting with the objects and characters in the story.
This is often achieved with the signer taking on the role of a referent
in the narrative and performing the actions (Metzger, 1995). CA is often
marked by certain non-manual cues such as the facial expression and the
breaking of the eye contact. This is to say, the narrator's head, face,
torso, arms, and as an optional marker, hands represent those of the
associated referent(s) (Smith and Cormier, 2014). As for the verbal
constructions, to identify the predicate types in ASL, Padden (1986)
suggested a triadic categorization of verbs in sign languages: plain
verbs, agreement verbs, and spatial verbs. According to this
classification, plain verbs (\ref{ex:9b}) do not agree with the subject
or the object. This means that they are not inflected for number or
person. This contrasts with agreeing verbs (\ref{ex:9c}) the starting or
the ending point of which can be modified to specific loci for person
and number agreement while spatial verbs (as in \textsc{ix book put} `I
placed the book there') agree with locative arguments but like plain
verbs, they are not inflected for number or person. TİD too licenses
null arguments. Following this classification, Sevinç (2006) argued that
although it is common in TİD to drop arguments, there were certain
limitations on the null marking of the arguments for agreement and plain
verbs.

@@

In both communication modalities (visual and auditory), children start
to produce more complex and coherent stories with appropriate and
informative linguistic forms to indicate anaphoric or cataphoric
references after age 10 (Hickmann et al., 1996; Rathmann et al., 2007).
Prior to this age, both signing and speaking children fail to refer to
entities consistently since their pragmatic abilities are yet to
develop. The manipulation of linguistic forms for discourse functions to
signal topic continuance or shift is difficult to master until the
beginning of formal education. These findings are also comparable across
different sign languages including BSL (Morgan, 2005), ASL (Loew, 1984),
and DGS (Becker, 2009), pointing out to the fact that the mastery of
reference tracking in narratives is one of the most protracted
achievements in the acquisition of signed discourse.

@@

Still, only very few research examined whether linguistically deprived
signers (i.e., deaf signers with delayed exposure to a sign language)
also display redundancy in their reference tracking. Becker (2009)
examined 6 DGS signing deaf children (3 native and 3 late)`s narrative
abilities. She reported that late signing children generated more
explicit forms such as nominals when referring to the characters in the
story compared to native learners, concluding that first language
deprivation results in an inadequate representation of the addressee and
less efficient use of morphosyntax. Comparably, Cormier et al. (2013)
investigated the use of CA in 15 BSL among deaf native and late adult
signers and found that late signers were redundant in topic continuity.
In other words, the late signers produced more explicit forms like
nominals, and fewer instances of null marking when maintaining previous
referents. Most relevantly to the present study, Gür (2018) studied the
development of narrative skills among 10 native and 10 late deaf
children and deaf adult signers of TİD. She found that native-signing
children and adults had more instances of explicit referent
introduction, set the scene more frequently, and employed CA in an
appropriate manner more than corresponding late-signing children and
adults. Although this study presented preliminary findings on age of
acquisition effects on narrative competence, it does not investigate how
native and late signers employ different linguistic tools based on the
accessibility of the referent. Therefore, to our knowledge, there is a
lack of research directly examining the effects of first language
deprivation on deaf adult signers' selection of REs in signed
narratives.

@@

Following Berman and Slobin (2013), we first identified the clause
boundaries in each sentence indicated by the presence of predicates and
informative non-manual markers (e.g., head nod). On the second tier, we
coded the discourse status for each referent. Having adapted the
classification of Gullberg (2006), we considered all referents mentioned
for the first time as introduced regardless of their syntactic function
in the sentence. Only subjects were coded for maintenance and
re-introduction. We coded subject referents as maintained only if they
had been referred to in the previous clause. To this end, the referents
which had been discontinued across one or more sentences were all
regarded re-introduced into the discourse. On a third tier, we had the
RE type for a single reference with the main categories nominal,
pronominal, null marking, and classifier. Under nominal, we identified
bare nouns, modified nouns, and fingerspelled nouns. For classifiers, we
distinguished among three types of constructions: WECL, BPCL, and ExtCL.
Finally, we had CA, agreement, plain tags for null constructions. We
were unable to observe any instance of HCLs occurring in isolation
(i.e., not part of CA). The annotation tiers are presented in Table
\ref{tab:tab-2}.

@@

\begin{table}
\small
\caption{Distribution of all discourse status by RE types}
\label{tab:my-table}
\begin{tabular}{llll}
\hline
                     & Introduced    & Maintained     & Re-introduced \\ \hline
\textit{Nominal:}    & 92.78\% (989) & 5.63\% (93)    & 63.65\% (907) \\
\textit{Pronominal:} & 0.09\% (1)    & 0.24\% (4)     & 2.46\% (35)   \\
\textit{Classifier:} & 5.35\% (57)   & 18.29\% (302)  & 8\% (114)     \\
\textit{Null:}       & 1.78\% (19)   & 75.83\% (1252) & 25.89\% (369) \\ \hline
\end{tabular}%
\end{table}

@@

\begin{table}
\small
\caption{Distribution of nominals by discourse status}
\label{tab:my-table2}
\begin{tabular}{lllll}
\hline
                        & Bare Noun  & Modified Noun & FS Noun \\ \hline
\textit{Introduced:}    & 48.93\% (842) & 53.56\% (139)     & 100\% (8) \\
\textit{Maintained:}    & 4.82\% (83)  & 3.85\% (10)      & 0\% (0) \\
\textit{Re-introduced:} & 46.25\% (796) & 42.69\% (111)     & 0\% (0) \\ \hline
\end{tabular}
\end{table}

\begin{table}
\small
\caption{Distribution of pronominals by discourse status}
\label{tab:my-table3}
\begin{tabular}{ll}
\hline
                        & Pronominal  \\ \hline
\textit{Introduced:}    & 2.5\% (1)   \\
\textit{Maintained:}    & 10\% (4)    \\
\textit{Re-introduced:} & 87.5\% (35) \\ \hline
\end{tabular}
\end{table}
